Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-31 ms.tex
fixed zeta discussion

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\begin{equation}
\dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T),
\label{eq:qdotcond}
 \end{equation}
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conductivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limiter $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

Hydrostatic equilibrium furthermore determines the value of the stagnation radius (Appendix \ref{app:rs}, eq.~\ref{eq:rs2main})
\begin{align}
  \rs&=\frac{G \Mbh}{\densSlope \vw^2}\left[\frac{13+ 8\Gamma}{4+2\Gamma}\right]\nonumber\\
  &=\frac{G \Mbh}{\densSlope (v_w^{2}+\sigma_0^2)}\left[4
    \frac{\Mstar|_{\rs}}{\Mbh} +\frac{13+
      8\Gamma}{4+2\Gamma}-\densSlope \frac{3}{2+\Gamma}\right]. 
\label{eq:stag_analytic}
\end{align}

If the stagnation radius is well inside the influence radius, $\rsoi$,
then $\Mstar/\Mbh|_{\rs}\ll1$. Also, equation~\eqref{eq:rsRinf}
implies that $v_w\gg \sigma_0$, since
$\rs/\rsoi=\frac{k(\Gamma)}{3\zeta^2}$, where
$\zeta=\sqrt{\frac{v_w^2+\sigma_0^2}{\sigma_0^2}}$ and $k(\Gamma)$ is
of order unity then equation~\eqref{eq:stag_analytic} simplifies to
\begin{align}
  &\rs \underset{v_w \gtrsim \sigma_0}\approx
  \left(\frac{13+8\Gamma}{4+2\Gamma}-\densSlope
    \frac{3}{2+\Gamma}\right)\frac{G \Mbh}{\densSlope v_w^2}\nonumber\\
  &\approx \begin{cases} 8
    \, \pc \,\, \Mbheight v_{500}^{-2}\, \pyear& \text{core} \\
    4 \, \pc \,\, \Mbheight v_{500}^{-2} \, \pyear & \text{cusp},
  \end{cases}
  \label{eq:stag_simple}
\end{align}
where we have used equation~\eqref{eq:densSlope} to estimate
$\densSlope$: $\densSlope\approx 1$ and $\densSlope \approx 0.6$ for
core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies.  This
expression is similar to that obtained by \citet{Volonteri+11} on more
heuristic grounds (their eq.~6).

In the opposite limit of weak heating ($v_w\lsim \sigma_0$) the
stagnation radius moves towards the break radius, $\rb$ in the stellar
density profile, formally implying that almost entire ISM of the
galaxy is accreting.  This result makes intuitive sense: gas is
supplied to the nuclear by stars which are gravitationally bound to
the black hole, so outflows are possible only if the specific heating
rate $\sim v_{w}^{2}$ significantly exceeds the specific gravitational
binding energy.

More precisely we find that there is a critical $\zeta_c$
($\zeta\equiv \sqrt{v_w^2+\sigma_0^2}/\sigma_0$), such that if
$\zeta\lsim\zeta_c$, then $\rs\simeq\rb$. For $\rb\gsim\rsoi$ we find

\begin{align}
\zeta_c\simeq \left(\frac{\rb}{\rsoi}\right)^{0.5(1-\Gamma)},
\end{align}

as described in Appendix~\ref{app:be}.

>>>>

To isolate the physics of greatest interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\begin{equation}
\dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T),
\label{eq:qdotcond}
 \end{equation}
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conductivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limiter $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

Hydrostatic equilibrium furthermore determines the value of the stagnation radius (Appendix \ref{app:rs}, eq.~\ref{eq:rs2main})
\begin{align}
  \rs&=\frac{G \Mbh}{\densSlope \vw^2}\left[\frac{13+ 8\Gamma}{4+2\Gamma}\right]\nonumber\\
  &=\frac{G \Mbh}{\densSlope (v_w^{2}+\sigma_0^2)}\left[4
    \frac{\Mstar|_{\rs}}{\Mbh} +\frac{13+
      8\Gamma}{4+2\Gamma}- \frac{3\nu}{2+\Gamma}\right]. 
\label{eq:stag_analytic}
\end{align}
For low heating rates $v_w \ll \sigma_0$, the stagnation radius resides well inside the SMBH sphere of influence.  In this case $\Mstar|_{\rs}/\Mbh \ll1$, such that equation~\eqref{eq:stag_analytic} simplifies to
\begin{align}
  &\rs \underset{v_w \gtrsim \sigma_0}\approx
  \left(\frac{13+8\Gamma}{4+2\Gamma}-
    \frac{3\nu}{2+\Gamma}\right)\frac{G \Mbh}{\nu v_w^2}\nonumber\\
  &\approx \begin{cases} 8
    \, \pc \,\, \Mbheight v_{500}^{-2}\, \pyear& \text{core} \\
    4 \, \pc \,\, \Mbheight v_{500}^{-2} \, \pyear & \text{cusp},
  \end{cases}
  \label{eq:stag_simple}
\end{align}
where we have used equation~\eqref{eq:densSlope} to estimate
$\densSlope$ separately for
core ($\Gamma = 0.1$; $\densSlope\approx 1$) and cusp ($\Gamma = 0.8$; $\densSlope \approx 0.6$) galaxies.  This
expression is similar to that obtained by \citet{Volonteri+11} on more
heuristic grounds (their eq.~6).

In the opposite limit of weak heating ($v_w\lsim \sigma_0$) the stagnation radius moves to large radii, approaching the break radius $\rb$ in the stellar density profile, formally implying that almost entire ISM of the galaxy is accreting.  In particular, in Appendix \ref{app:be} we show that $\rs \Rightarrow \rb$ for heating below a critical threshold
\be
\zeta\equiv \frac{\sqrt{v_w^2+\sigma_0^2}}{\sigma_0} < \zeta_c \approx \left(\frac{\rb}{\rsoi}\right)^{0.5(1-\Gamma)}.
\label{eq:zetacrit}
\ee
This result makes intuitive sense: gas is supplied to the nuclear by stars which are gravitationally bound to the black hole, so outflows are possible only if the specific heating rate $\sim v_{w}^{2}$ significantly exceeds the specific gravitational binding energy.  Condition (\ref{eq:zetacrit}) approximately corresponds to the condition that the wind heating parameter be less than the escape speed from the break radius, $v_{\rm esc} \sim (GM_{\star}|_{r_{\rm b}}/r_b)^{1/2}$, where $M_{\star} \propto r^{2-\Gamma}$.
